---
title: "Exploring biotaxa"
author: "Hsun-Yi Hsieh"
date: "5/30/2018"
output: html_document
runtime: shiny
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This page evaluates the relationship between predicted and actual species richness of well-studiex phyla in Animalia.

First, explore the most frequently recorded phyla in Animalia
```{r, echo = TRUE, eval = TRUE}
library(shiny)
library(biotaxa)
library(dplyr)
topphyla <- topn("Animalia", "Phylum", 12)
topphyla
```

Take looks of their discovery curves and how well logistic regressions fit.
```{r, echo = TRUE}
library(biotaxa)

taxamodel("Arthropoda", "Species", "logistic")
taxamodel("Mollusca", "Species", "logistic")
taxamodel("Chordata", "Species", "logistic")
taxamodel("Cnidaria", "Species", "logistic")
taxamodel("Echinodermata", "Species", "logistic")
taxamodel("Annelida", "Species", "logistic")
taxamodel("Bryozoa", "Species", "logistic")
taxamodel("Porifera", "Species", "logistic")
taxamodel("Nematoda", "Species", "logistic")
taxamodel("Brachiopoda", "Species", "logistic")
taxamodel("Nemertea", "Species", "logistic")
taxamodel("Platyhelminthes", "Species", "logistic")

```

add the most frequently-studied classes of Animalia.
find the ones with highest species richness.
```{r}
topclasses <- topn("Animalia", "Class", 12)
topclasses
```
```{r, echo = TRUE}
taxamodel("Malacostraca", "Species", "logistic")
taxamodel("Gastropoda", "Species", "logistic")
taxamodel("Actinopterygii", "Species", "logistic")
taxamodel("Polychaeta", "Species", "logistic")
taxamodel("Gymnolaemata", "Species", "logistic")
taxamodel("Bivalvia", "Species", "logistic")
taxamodel("Demospongiae", "Species", "logistic")
taxamodel("Hexanauplia", "Species", "logistic")
taxamodel("Anthozoa", "Species", "logistic")
taxamodel("Hydrozoa", "Species", "logistic")
```

##Collect the predictied and actual values in one data frame
```{r, echo = TRUE, eval = TRUE}
library(stringr)
library(drc)
library(data.table)
taxa <- c("Arthropoda","Mollusca", "Chordata", "Cnidaria", "Echinodermata","Annelida", "Bryozoa", "Porifera", "Nematoda", "Brachiopoda", "Nemertea", "Platyhelminthes")

ranks <- rep("Species", length(taxa))
strLines <- mapply(taxa_rich, taxa, ranks)
prediction <- function(strLine) {
  str_extract_all(strLine, "\\(?[0-9,.]+\\)?")[[1]][1]
}
preds <- as.numeric(unlist(lapply(strLines, prediction)))
actual <- function(strLine) {
  str_extract_all(strLine, "\\(?[0-9,.]+\\)?")[[1]][3]
}
actuals <- as.numeric(unlist(lapply(strLines, actual)))
ndf <- data.frame(taxa, actuals, preds)

#dev.new(width=5, height=5)
x11(width = 5, height = 5)
plot(actuals, preds, xlab = "Predicted number of species", ylab = "Actual number of species", pch = 19)
abline(0,1, lty = 2)
#text(actuals~preds, labels = taxa, data = ndf, cex = 0.7, font = 1.5)
```

add points of classes in Animalia to the plot
```{r, echo = TRUE, eval = TRUE}
taxa <- c("Malacostraca", "Gastropoda", "Actinopterygii", "Polychaeta", "Bivalvia", "Demospongiae", "Hexanauplia", "Anthozoa", "Hydrozoa")
ranks <- rep("Species", length(taxa))
strLines <- mapply(taxa_rich, taxa, ranks)

prediction_class <- function(strLine) {
  str_extract_all(strLine, "\\(?[0-9,.]+\\)?")[[1]][1]
}
preds_classes <- unlist(lapply(strLines, prediction_class))

actual_class <- function(strLine) {
  str_extract_all(strLine, "\\(?[0-9,.]+\\)?")[[1]][3]
}
actuals_classes <- unlist(lapply(strLines, actual_class)) 

ndf_classes <- data.frame(actuals_classes, preds_classes)
x11(width=5, height=5)
#windows.options(width = 5, height = 5)
plot(actuals, preds, xlab = "Predicted number of species", ylab = "Actual number of species", pch = 19)
abline(0,1, lty = 2)
points(preds_classes, actuals_classes, pch = 2, cex = 0.8)
```

Now collect all species richness of all subordinate taxa in *Animalia*.
```{r}
data_m <- subset(data_m, Kingdoms == "Animalia")
dim(data_m)
richness <- function(rank) {
  length(spetaxalist(data_m, rank))
}

ranks <- c("Phylum", "Class", "Order", "Family", "Genus")

allrichness <- unlist(lapply(ranks, richness))
forx <- seq(1:length(allrichness))
model <- lm(log(log(allrichness)) ~ forx)
summary(model)

plot(log(log(allrichness)) ~ forx, ylab = "log10(log10(species richness))", xlab = "", pch = 19, ylim = c(1, 2.5))
abline(model, col = "red", lwd = 3, lty = 2)
title("Species richness of subordinate taxa in Animalia")
text(1.2, 1.4, "Phylum", cex = 0.8)
text(2.2, 1.6, "Class", cex = 0.8)
text(3.2, 1.8, "Order", cex = 0.8)
text(4.2, 1.9, "Family", cex = 0.8)
text(4.9, 2.0, "Genus", cex = 0.8)

```

