---
title: "exploring power law relationships between hierarchical taxonomic categories"
author: "Hsun-Yi Hsieh"
date: "6/5/2018"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Note: users are encouraged to appropriately use survey data while employing the code for investigating the power law relationship between taxonomic categories. </br>

The method is adopted from **Passy and Legendre, Power law relationships among hierarchical categories in algae reveal a new paradox of the plankton, Global Ecology and Biogeography (2006) 15, 528-535**, along with other studies of the kind.

The dataset should include hierarhical categories of species. Here the example dataset is only for demonstration, rather than a basis for the generation of scientific perspectives. 

Data should be composed of, for instance, *samples of phytoplankton, samples of trees, etc.*
```{r}
df <- subset(data_m, Phyla == "Mollusca")
n_classes <- length(unique(df$Classes))
n_orders <- length(unique(df$Order))
n_families <- length(unique(df$Families))
n_genera <- length(unique(df$Genera))
n_species <- length(unique(df$AphiaIDs))
```

randomization
``{r}

```

